A Closer Look at Network Flow on Evolving Graphs

### 2018-05-15 paper.tex
update temporal katz and temporal resolvent betweenness section

\title{A Closer Look at Network Flow on Evolving Graph Centrality}

@@

% what is centrality?
Centrality measures the importance of nodes within a graph.
% current evolving graph centrality approach
To generalize centrality measures for evolving graphs, there are two common approaches.
First, we could exploit the fact that matrix product of adjacency matrices at different time stamps counts the number of dynamic walks in an evolving graph. Using this idea, we can generalize walk-based centrality measures such as
the Katz Centrality \cite{estrada09} and Communicability Betweenness \cite{alsayed15}.
Second, we could generalize the definition of shortest paths as shortest temporal paths and replace shortest paths in centrality measures with shortest temporal paths. This way we can define temporal betweenness centrality and temporal closeness centrality \cite{nicosia13}.

% our approach
We observe both approaches ignore the communication time between edges at different time stamps, i.e., the time gap between different graph slices. We take account of this communication time by deriving centrality measures from original ideas and show communication time has important impact on the final rating (and ranking).
We note that our goal is not about providing faster centrality algorithms which is out of the scope of this paper.

@@

\begin{lstlisting}[caption={Temporal Katz Centrality of Single Node},label={lst:katz}]
function temporal_katz(g::AbstractEvolvingGraph,
  start::TimeNode; alpha = 0.2, max_level = 10)
    score = 0.
    v = start
    fronter = [v]
    level = 0
    while level < max_level
        next = []
        for u in fronter
            for v in forward_neighbors(g, u)
                push!(next, v)
                td = temporal_distance(start, u)
                d = td + level
                score += alpha^d
            end
        end
        fronter = next
        level += 1
    end
    return score
end
\end{lstlisting}

% explain the code
The algorithm \texttt{temporal\_katz} performs breadth first search (BFS) starting at \texttt{TimeNode} $v$, which represent a node at a specific time stamp \footnote{See also \url{https://etymoio.github.io/EvolvingGraphs.jl/latest/base.html#EvolvingGraphs.TimeNode}}.
For each \texttt{TimeNode} $v$, we accumulate the influence of linked nodes according to their spatial and temporal
distance from $v$. The spatial distance is recorded in variable \texttt{level} and the temporal distance is calculated by
\texttt{temporal\_distance}. The algorithm stops searching when \texttt{level} is larger or equal to \texttt{max\_level}.
We set \texttt{alpha} to be $0.2$ and \texttt{max\_level} to be $10$. Then
% example on 3 timestamp evolving graph
for Figure \ref{fig:katz_eg2}, we have
\begin{lstlisting}
Dict{Any,Any} with 7 entries:
  TimeNode(B, 2) => 1.0
  TimeNode(B, 3) => 0.0
  TimeNode(A, 2) => 3.58333
  TimeNode(C, 1) => 1.06867
  TimeNode(A, 3) => 1.0
  TimeNode(B, 1) => 3.25373
  TimeNode(C, 2) => 1.71667
\end{lstlisting}
The total scores at all time stamps are
\begin{lstlisting}
DataStructures.DefaultDict{Any,Any,Float64} with 3 entries:
  "B" => 4.25373
  "A" => 4.58333
  "C" => 2.78533
\end{lstlisting}
Notice that $A$ is still the most important node in the network. However, now $B$ is considered more important
than $C$. This is because we consider causal edges as well as static edges. Node $B$ is active at all three time stamps
and more connected causally than node $C$.

For Figure \ref{fig:katz_eg3}, we have
\begin{lstlisting}
Dict{Any,Any} with 7 entries:
  TimeNode(C, 1)   => 1.06833
  TimeNode(B, 100) => 0.0
  TimeNode(C, 2)   => 1.70833
  TimeNode(B, 2)   => 1.0
  TimeNode(B, 1)   => 3.25367
  TimeNode(A, 100) => 1.0
  TimeNode(A, 2)   => 3.54167
\end{lstlisting}
The sum of node scores at each time stamp is
\begin{lstlisting}
DataStructures.DefaultDict{Any,Any,Float64} with 3 entries:
  "B" => 4.25367
  "A" => 4.54167
  "C" => 2.77667
\end{lstlisting}
We see the scores of all three nodes are decreased when the last time stamp is $100$.
Node $A$ is affected the most.

@@

Betweenness centrality measures the importance of a node as the ability to facilitate the communication between other nodes in the network. Commonly it is defined based on the shortest paths. However, key messages do not necessarily follow the shortest paths and so it makes senses to consider walks instead.
Recall the $(i,j)$ entry of the resolvent matrix $(I - \alpha A)^{-1}$ provide information about the communication from $i$ to $j$. We could therefore define the resolvent betweenness for node $v$ as
\begin{equation}
  \sum_{i}\sum_j \frac{(I-\alpha A)^{-1}_{i,j} - (I - \alpha (A - E_v))^{-1}_{i,j}}{(I - \alpha A)^{-1}_{i,j}}, \quad i \ne j, i \ne v, j \ne v.
\end{equation}
where $E_v$ has non-zeros only in row and column $v$, and in the $v$th row and column has $1$ where $A$ has $1$.
The matrix $A - E_i$ represents a graph with all edges involving the node $v$ removed.
For $(I-\alpha A)^{-1}_{i,j}$ we could simply modify Listing \ref{lst:katz} so that it only accumulate scores when the end node is $j$. For the $(I - \alpha (A-E_v))^{-1}_{i,j}$ part, we can apply the above the same algorithm on a graph with node $v$ and related edges removed.
\begin{lstlisting}
function temporal_resolvent_betweenness(g1::EvolvingGraph,
  g2::EvolvingGraph, v::TimeNode; alpha = 0.2, k = 10)
    r = 0.
    ns = active_nodes(g1)
    for i in ns
        for j in ns
            if i != j && i != v && j != v
                score1 = temporal_katz(g1, i, j, alpha = alpha, k = k)
                score2 = temporal_katz(g2, i, j, alpha = alpha, k = k)
                r += score1 == 0.0 ? 0 : (score1 - score2)/score1
            end
        end
    end
    return r
end
\end{lstlisting}
For example, by removing node $C$ a time stamp $2$ from Figure \ref{fig:katz_eg2} we find
the temporal resolvent betweenness score is $6.09$. While the temporal resolvent betweenness score
of node $A$ at time stamp $3$ is $5.0$. This shows $(C,2)$ is more important than $(A,3)$.

Now consider Figure \ref{fig:katz_eg3}. If we remove $(C,2)$, the temporal resolvent betweenness score is $4.20$.
The analysis for temporal communicability betweenness centrality is similar.

@@

>>>>

\title{A Closer Look at Network Flow on Evolving Graphs}

@@

% what is centrality?
Centrality measures the importance of nodes within a graph.
% current evolving graph centrality approach
To generalize centrality measures for evolving graphs there are two common approaches.
First, we could exploit the fact that matrix product of adjacency matrices at different time stamps counts the number of dynamic walks in an evolving graph. Using this idea, we can generalize walk-based centrality measures such as
the Katz Centrality \cite{estrada09} and Communicability Betweenness \cite{alsayed15}.
Second, we could generalize the definition of shortest paths as shortest temporal paths and replace shortest paths in centrality measures with shortest temporal paths where possible. This way we can define temporal betweenness centrality and temporal closeness centrality \cite{nicosia13}.

% our approach
We observe both approaches ignore the communication time between edges at different time stamps, i.e., the time gap between different (static) graph slices. We take account of this communication time by deriving centrality measures from the original theses of the ideas and show communication time has important impact on the final node ratings (and rankings).
We note that our goal is not about providing efficient centrality algorithms which is out of the scope of this paper.

@@

\begin{lstlisting}[caption={Temporal Katz Centrality of Single Node},label={lst:katz}]
function temporal_katz(g::AbstractEvolvingGraph,
  start::TimeNode; alpha = 0.2, max_level = 10)
    score = 0.
    v = start
    fronter = [v]
    level = 0
    while level < max_level
        next = []
        for u in fronter
            for v in forward_neighbors(g, u)
                push!(next, v)
                if node_key(v) != node_key(u)
                  td = temporal_distance(start, u)
                  d = td + level
                  score += alpha^d
                end
            end
        end
        fronter = next
        level += 1
    end
    return score / num_edges(g)
end
\end{lstlisting}

% explain the code
The algorithm \texttt{temporal\_katz} performs a breadth first search (BFS) on evolving graph $g$ that
starts at \texttt{TimeNode} $v$, which represents a node at a specific time stamp \footnote{See also \url{https://etymoio.github.io/EvolvingGraphs.jl/latest/base.html#EvolvingGraphs.TimeNode}}.
For each \texttt{TimeNode} $v$, we accumulate the influence of linked nodes according to their spatial and temporal
distance from $v$. The spatial distance is recorded in variable \texttt{level} and the temporal distance is calculated by
\texttt{temporal\_distance}. The algorithm stops searching when \texttt{level} is larger or equal to \texttt{max\_level}.
We set \texttt{alpha} to be $0.2$ and \texttt{max\_level} to be $10$. Then
% example on 3 timestamp evolving graph
for Figure \ref{fig:katz_eg2}, we have
\begin{lstlisting}
  TimeNode(A, 2) => 0.466667
  TimeNode(A, 3) => 0.2
  TimeNode(B, 2) => 0.0
  TimeNode(B, 3) => 0.0
  TimeNode(B, 1) => 0.202347
  TimeNode(C, 1) => 0.0117333
  TimeNode(C, 2) => 0.293333
\end{lstlisting}
The total scores at all time stamps are
\begin{lstlisting}
  "A" => 0.666667
  "B" => 0.202347
  "C" => 0.305067
\end{lstlisting}
Notice that $A$ is the most important node in the evolving graph and $C$ is the second most important node.
Thus the ranking agrees with the Katz centrality rankings in Section \ref{sec:motivation}.

For Figure \ref{fig:katz_eg3}, we have
\begin{lstlisting}
  TimeNode(A, 10)  => 0.458333
  TimeNode(A, 100) => 0.2
  TimeNode(B, 1)   => 0.2
  TimeNode(B, 10)  => 0.0
  TimeNode(B, 100) => 0.0
  TimeNode(C, 1)   => 2.98666e-8
  TimeNode(C, 10)  => 0.291667
\end{lstlisting}
The sum of node scores at each time stamp is
\begin{lstlisting}
  "A" => 0.658333
  "B" => 0.2
  "C" => 0.291667
\end{lstlisting}
We see the scores of all three nodes are decreased when it takes longer to communicate between different time stamps. In this case, the overall ranking is not affected but in general the communication time between time stamps can change the overall ranking.

@@

Betweenness centrality measures the importance of a node as the ability to facilitate the communication among other nodes in the network. Commonly it is defined based on the shortest paths. However, key messages do not necessarily follow the shortest paths and thus it makes senses to consider walks instead of shortest paths.
Recall the $(i,j)$ entry of the resolvent matrix $(I - \alpha A)^{-1}$ provide information about the communication from $i$ to $j$. We could therefore define the resolvent betweenness for node $v$ as
\begin{equation}
  \sum_{i}\sum_j \frac{(I-\alpha A)^{-1}_{i,j} - (I - \alpha (A - E_v))^{-1}_{i,j}}{(I - \alpha A)^{-1}_{i,j}}, \quad i \ne j, i \ne v, j \ne v.
\end{equation}
where $E_v$ has non-zeros only in row and column $v$, and in the $v$th row and column has $1$ where $A$ has $1$.
The matrix $A - E_i$ represents a graph with all edges involving the node $v$ removed.
For $(I-\alpha A)^{-1}_{i,j}$ we could simply modify Listing \ref{lst:katz} so that it only accumulate scores when the end node is $j$. For the $(I - \alpha (A-E_v))^{-1}_{i,j}$ part, we can apply the above the same algorithm on a graph with node $v$ and related edges removed.
Therefore the general temporal resolvent betweenness centrality can be defined as
\begin{lstlisting}
function temporal_resolvent_betweenness(g1, g2,
  v::TimeNode; alpha = 0.2, k = 10)
    r = 0.
    ns = active_nodes(g1)
    for i in ns
        for j in ns
            if i != j && i != v && j != v
                score1 = temporal_katz(g1, i, j,
                    alpha = alpha, k = k)
                score2 = temporal_katz(g2, i, j,
                    alpha = alpha, k = k)
                r += score1 == 0.0 ? 0 :
                    (score1 - score2)/score1
            end
        end
    end
    return r
end
\end{lstlisting}
Here \texttt{temporal\_katz} only takes account of walks from node $i$ to $j$. Evolving graph
\texttt{g1} is the original evolving graph and evolving graph \texttt{g2} is \texttt{g1} with all the edges
related to \texttt{v} removed.

For example, by removing node $(A, 2)$, i.e., node $A$ at time stamp $2$, from Figure \ref{fig:katz_eg2} we find
the temporal resolvent betweenness score of node $A$ is $5.0$. While the temporal resolvent betweenness score of node $(B,3)$ is $4.0$. This shows that $(A,2)$ is more important than $(B,3)$.
We note that temporal communicability betweenness centrality can be analysed similarly.

@@

One potential problem is the computational cost. Since the dimension of the block matrix
is $|V| \times |T|$, where $|V|$ represents the number of nodes and $|T|$ represents the number of time stamps. The block matrix may become very large for an evolving graph with many time stamps.
